---
title: "Hands-on Exercise 9"
author: "Marga Thura"
date: "June 19 2025"
date-modified: "June 19 2025"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# 13 Creating Ternary Plot with R

13.1 Overview

Ternary plots are a way of displaying the distribution and variability of three-part compositional data. (For example, the proportion of aged, economy active and young population or sand, silt, and clay in soil.) It’s display is a triangle with sides scaled from 0 to 1. Each side represents one of the three components. A point is plotted so that a line drawn perpendicular from the point to each leg of the triangle intersect at the component values of the point.

The hands-on exercise consists of four steps:

-   Install and launch **tidyverse** and **ggtern** packages.

-   Derive three new measures using *mutate()* function of **dplyr** package.

-   Build a static ternary plot using *ggtern()* function of **ggtern** package.

-   Build an interactive ternary plot using *plot-ly()* function of **Plotly R** package.

## 13.2 Installing and launching R packages

-   [**ggtern**](http://www.ggtern.com/), a ggplot extension specially designed to plot ternary diagrams. The package will be used to plot static ternary plots.

-   [**Plotly R**](https://plot.ly/r/), an R package for creating interactive web-based graphs via plotly’s JavaScript graphing library, plotly.js . The **plotly R** libary contains the *ggplotly* function, which will convert **ggplot2** figures into a Plotly object.

```{r}
pacman::p_load(plotly, ggtern, tidyverse)
```

## 13.3 Data Preparation

### 13.3.1 The data

### 13.3.2 Importing Data

To important *respopagsex2000to2018_tidy.csv* into R, ***read_csv()*** function of **readr** package will be used.

```{r}
pop_data <- read_csv("data/respopagsex2000to2018_tidy.csv") 
```

### 13.3.3 Preparing the Data

Next, use the ***mutate()*** function of **dplyr** package to derive three new measures, namely: young, active, and old.

```{r}
agpop_mutated <- pop_data %>%
  mutate(`Year` = as.character(Year))%>%
  spread(AG, Population) %>%
  mutate(YOUNG = rowSums(.[4:8]))%>%
  mutate(ACTIVE = rowSums(.[9:16]))  %>%
  mutate(OLD = rowSums(.[17:21])) %>%
  mutate(TOTAL = rowSums(.[22:24])) %>%
  filter(Year == 2018)%>%
  filter(TOTAL > 0)
```

## 13.4 Plotting Ternary Diagram with R

### 13.4.1 4.1 Plotting a static ternary diagram

Use ***ggtern()*** function of **ggtern** package to create a simple ternary plot.

```{r}
ggtern(data=agpop_mutated,aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point()
```

```{r}
ggtern(data=agpop_mutated, aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point() +
  labs(title="Population structure, 2015") +
  theme_rgbw()
```

### 13.4.2 Plotting an interative ternary diagram

The code below create an interactive ternary plot using ***plot_ly()*** function of **Plotly R**.

```{r}
label <- function(txt) {
  list(
    text = txt, 
    x = 0.1, y = 1,
    ax = 0, ay = 0,
    xref = "paper", yref = "paper", 
    align = "center",
    font = list(family = "serif", size = 15, color = "white"),
    bgcolor = "#b3b3b3", bordercolor = "black", borderwidth = 2
  )
}

# reusable function for axis formatting
axis <- function(txt) {
  list(
    title = txt, tickformat = ".0%", tickfont = list(size = 10)
  )
}

ternaryAxes <- list(
  aaxis = axis("Young"), 
  baxis = axis("Active"), 
  caxis = axis("Old")
)

# Initiating a plotly visualization 
plot_ly(
  agpop_mutated, 
  a = ~YOUNG, 
  b = ~ACTIVE, 
  c = ~OLD, 
  color = I("black"), 
  type = "scatterternary"
) %>%
  layout(
    annotations = label("Ternary Markers"), 
    ternary = ternaryAxes
  )
```

# 6 Visual Correlation Analysis

# 6.1 Overview

In this hands-on exercise, we will learn how to plot data visualisation for visualising correlation matrix with R. It consists of three main sections. First, we will learn how to create correlation matrix using [*pairs()*](https://www.rdocumentation.org/packages/graphics/versions/3.6.0/topics/pairs) of R Graphics. Next, we will learn how to plot corrgram using **corrplot** package of R. Lastly, we will learn how to create an interactive correlation matrix using plotly R.

## **6.2 Installing and Launching R Packages**

```{r}
pacman::p_load(corrplot, ggstatsplot, tidyverse)
```

## **6.3 Importing and Preparing The Data Set**

In this hands-on exercise, the [Wine Quality Data Set](https://archive.ics.uci.edu/ml/datasets/wine+quality) of UCI Machine Learning Repository will be used. The data set consists of 13 variables and 6497 observations.

### **6.3.1 Importing Data**

```{r}
wine <- read_csv("data/wine_quality.csv")
```

## **6.4 Building Correlation Matrix: *pairs()* method**

### **6.4.1 Building a basic correlation matrix**

```{r}
pairs(wine[,1:11])
```

```{r}
pairs(wine[,2:12])
```

Figure below shows the scatter plot matrix of Wine Quality Data. It is a 11 by 11 matrix.

### **6.4.2 Drawing the lower corner**

```{r}
pairs(wine[,2:12], upper.panel = NULL)
```

```{r}
pairs(wine[,2:12], lower.panel = NULL)
```

### **6.4.3 Including with correlation coefficients**

To show the correlation coefficient of each pair of variables instead of a scatter plot, [*panel.cor*](https://www.rdocumentation.org/packages/xcms/versions/1.48.0/topics/panel.cor) function will be used. This will also show higher correlations in a larger font.

```{r}
panel.cor <- function(x, y, digits=2, prefix="", cex.cor, ...) {
usr <- par("usr")
on.exit(par(usr))
par(usr = c(0, 1, 0, 1))
r <- abs(cor(x, y, use="complete.obs"))
txt <- format(c(r, 0.123456789), digits=digits)[1]
txt <- paste(prefix, txt, sep="")
if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
text(0.5, 0.5, txt, cex = cex.cor * (1 + r) / 2)
}

pairs(wine[,2:12], 
      upper.panel = panel.cor)
```

## **6.5 Visualising Correlation Matrix: ggcormat()**

### **6.5.1 The basic plot**

On of the advantage of using *ggcorrmat()* over many other methods to visualise a correlation matrix is it’s ability to provide a comprehensive and yet professional statistical report as shown in the figure below.

```{r}
ggstatsplot::ggcorrmat(
  data = wine, 
  cor.vars = 1:11)
```

```{r}
ggstatsplot::ggcorrmat(
  data = wine, 
  cor.vars = 1:11,
  ggcorrplot.args = list(outline.color = "black", 
                         hc.order = TRUE,
                         tl.cex = 10),
  title    = "Correlogram for wine dataset",
  subtitle = "Four pairs are no significant at p < 0.05"
)
```

```{r}
ggplot.component = list(
    theme(text=element_text(size=5),
      axis.text.x = element_text(size = 8),
      axis.text.y = element_text(size = 8)))
```

## **6.6 Building multiple plots**

```{r}
grouped_ggcorrmat(
  data = wine,
  cor.vars = 1:11,
  grouping.var = type,
  type = "robust",
  p.adjust.method = "holm",
  plotgrid.args = list(ncol = 2),
  ggcorrplot.args = list(outline.color = "black", 
                         hc.order = TRUE,
                         tl.cex = 10),
  annotation.args = list(
    tag_levels = "a",
    title = "Correlogram for wine dataset",
    subtitle = "The measures are: alcohol, sulphates, fixed acidity, citric acid, chlorides, residual sugar, density, free sulfur dioxide and volatile acidity",
    caption = "Dataset: UCI Machine Learning Repository"
  )
)
```

## **6.7 Visualising Correlation Matrix using corrplot Package**

### **6.7.1 Getting started with corrplot**

Before we can plot a corrgram using *corrplot()*, we need to compute the correlation matrix of wine data frame.

In the code chunk below, [*cor()*](https://www.rdocumentation.org/packages/stats/versions/3.6.0/topics/cor) of R Stats is used to compute the correlation matrix of wine data frame.

```{r}
wine.cor <- cor(wine[, 1:11])
```

Next, corrplot() is used to plot the corrgram by using all the default setting

```{r}
corrplot(wine.cor)
```

### **6.7.2 Working with visual geometrics**

In **corrplot** package, there are seven visual geometrics (parameter method) can be used to encode the attribute values. They are: circle, square, ellipse, number, shade, color and pie. The default is circle. As shown in the previous section, the default visual geometric of corrplot matrix is circle. However, this default setting can be changed by using the *method* argument as shown in the code chunk below.

```{r}
corrplot(wine.cor, 
         method = "ellipse") 
```

### **6.7.3 Working with layout**

*corrplor()* supports three layout types, namely: “full”, “upper” or “lower”. The default is “full” which display full matrix. The default setting can be changed by using the *type* argument of *corrplot()*.

```{r}
corrplot(wine.cor, 
         method = "ellipse", 
         type="lower")
```

```{r}
corrplot(wine.cor, 
         method = "ellipse", 
         type="lower",
         diag = FALSE,
         tl.col = "black")
```

### **6.7.4 Working with mixed layout**

With **corrplot** package, it is possible to design corrgram with mixed visual matrix of one half and numerical matrix on the other half. In order to create a coorgram with mixed layout, the [*corrplot.mixed()*](https://www.rdocumentation.org/packages/corrplot/versions/0.84/topics/corrplot.mixed), a wrapped function for mixed visualisation style will be used.

```{r}
corrplot.mixed(wine.cor, 
               lower = "ellipse", 
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black")
```

```{r}
corrplot.mixed(wine.cor, 
               lower = "ellipse", 
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black")
```

### **6.7.5 Combining corrgram with the significant test**

With corrplot package, we can use the *cor.mtest()* to compute the p-values and confidence interval for each pair of variables.

```{r}
wine.sig = cor.mtest(wine.cor, conf.level= .95)
```

We can then use the *p.mat* argument of *corrplot* function as shown in the code chunk below.

```{r}
corrplot(wine.cor,
         method = "number",
         type = "lower",
         diag = FALSE,
         tl.col = "black",
         tl.srt = 45,
         p.mat = wine.sig$p,
         sig.level = .05)
```

### **6.7.6 Reorder a corrgram**

Matrix reorder is very important for mining the hiden structure and pattern in a corrgram. By default, the order of attributes of a corrgram is sorted according to the correlation matrix (i.e. “original”). The default setting can be over-write by using the *order* argument of *corrplot()*. Currently, **corrplot** package support four sorting methods, they are:

-   “AOE” is for the angular order of the eigenvectors. See Michael Friendly (2002) for details.

-   “FPC” for the first principal component order.

-   “hclust” for hierarchical clustering order, and “hclust.method” for the agglomeration method to be used.

    -   “hclust.method” should be one of “ward”, “single”, “complete”, “average”, “mcquitty”, “median” or “centroid”.

-   “alphabet” for alphabetical order.

“AOE”, “FPC”, “hclust”, “alphabet”. More algorithms can be found in **seriation** package.

```{r}
corrplot.mixed(wine.cor, 
               lower = "ellipse", 
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               order="AOE",
               tl.col = "black")
```

**6.7.7 Reordering a correlation matrix using hclust**

If using **hclust**, ***corrplot()*** can draw rectangles around the corrgram based on the results of hierarchical clustering.

```{r}
corrplot(wine.cor, 
         method = "ellipse", 
         tl.pos = "lt",
         tl.col = "black",
         order="hclust",
         hclust.method = "ward.D",
         addrect = 3)
```

# 14 Heatmap for Visualising and Analysing Multivariate Data

In this hands-on exercise, we will gain hands-on experience on using R to plot static and interactive heatmap for visualising and analysing multivariate data.

## 14.2 Installing and Launching R Packages

In this hands-on exercise, the data of [World Happines 2018 report](https://worldhappiness.report/ed/2018/) will be used.

```{r}
pacman::p_load(seriation, dendextend, heatmaply, tidyverse)
```

## **14.3 Importing and Preparing The Data Set**

```{r}
wh <- read_csv("data/WHData-2018.csv")
```

### **14.3.2 Preparing the data**

```{r}
row.names(wh) <- wh$Country
```

### **14.3.3 Transforming the data frame into a matrix**

```{r}
wh1 <- dplyr::select(wh, c(3, 7:12))
wh_matrix <- data.matrix(wh)
```

## **14.4 Static Heatmap**

### **14.4.1 heatmap() of R Stats**

In this sub-section, we will plot a heatmap by using *heatmap()* of Base Stats. The code chunk is given below.

```{r}
wh_heatmap <- heatmap(wh_matrix,
                      Rowv=NA, Colv=NA)
```

***Note:***

-   By default, **heatmap()** plots a cluster heatmap. The arguments ***Rowv=NA*** and ***Colv=NA*** are used to switch off the option of plotting the row and column dendrograms.

To plot a cluster heatmap, we just have to use the default as shown in the code chunk below.

```{r}
wh_heatmap <- heatmap(wh_matrix)
```

The code chunk below normalises the matrix column-wise.\

```{r}
wh_heatmap <- heatmap(wh_matrix,
                      scale="column",
                      cexRow = 0.6, 
                      cexCol = 0.8,
                      margins = c(10, 4))
```

## **14.5 Creating Interactive Heatmap**

[**heatmaply**](http://talgalili.github.io/heatmaply/index.html) is an R package for building interactive cluster heatmap that can be shared online as a stand-alone HTML file.

### **14.5.1 Working with heatmaply**

The code chunk below shows the basic syntax needed to create n interactive heatmap by using **heatmaply** package.

```{r}
heatmaply(mtcars)
```

The code chunk below shows the basic syntax needed to create n interactive heatmap by using heatmaply package.

```{r}
heatmaply(wh_matrix[, -c(1, 2, 4, 5)])
```

### **14.5.2 Data trasformation**

When analysing multivariate data set, it is very common that the variables in the data sets includes values that reflect different types of measurement. In general, these variables’ values have their own range. In order to ensure that all the variables have comparable values, data transformation are commonly used before clustering.

Three main data transformation methods are supported by *heatmaply()*, namely: scale, normalise and percentilse.

#### 14.5.2.1 Scaling method

```{r}
heatmaply(wh_matrix[, -c(1, 2, 4, 5)],
          scale = "column")
```

#### 14.5.2.2 Normalising method

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]))
```

#### 14.5.2.3 Percentising method

```{r}
heatmaply(percentize(wh_matrix[, -c(1, 2, 4, 5)]))
```

### **14.5.3 Clustering algorithm**

### **14.5.4 Manual approach**

In the code chunk below, the heatmap is plotted by using hierachical clustering algorithm with “Euclidean distance” and “ward.D” method.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          dist_method = "euclidean",
          hclust_method = "ward.D")
```

### **14.5.5 Statistical approach**

In order to determine the best clustering method and number of cluster the *dend_expend()* and *find_k()* functions of **dendextend** package will be used.

First, the *dend_expend()* will be used to determine the recommended clustering method to be used.

```{r}
wh_d <- dist(normalize(wh_matrix[, -c(1, 2, 4, 5)]), method = "euclidean")
dend_expend(wh_d)[[3]]
```

Next, *find_k()* is used to determine the optimal number of cluster.

```{r}
wh_clust <- hclust(wh_d, method = "average")
num_k <- find_k(wh_clust)
plot(num_k)
```

With reference to the statistical analysis results, we can prepare the code chunk as shown below.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          dist_method = "euclidean",
          hclust_method = "average",
          k_row = 3)
```

### **14.5.6 Seriation**

## OLO

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "OLO")
```

## GW

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "GW")
```

## Mean

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "mean")
```

## None

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "none")
```

### **14.5.7 Working with colour palettes**

The default colour palette uses by **heatmaply** is *viridis*. heatmaply users, however, can use other colour palettes in order to improve the aestheticness and visual friendliness of the heatmap.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "none",
          colors = Blues)
```

In the code chunk below, the Blues colour palette of rColorBrewer is used

### **14.5.8 The finishing touch**

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          Colv=NA,
          seriate = "none",
          colors = Blues,
          k_row = 5,
          margins = c(NA,200,60,NA),
          fontsize_row = 4,
          fontsize_col = 5,
          main="World Happiness Score and Variables by Country, 2018 \nDataTransformation using Normalise Method",
          xlab = "World Happiness Indicators",
          ylab = "World Countries"
          )
```

# 15 Visual Multivariate Analysis with Parallel Coordinates Plot

## **15.1 Overview**

By the end of this hands-on exercise, we will gain hands-on experience on:

-   plotting statistic parallel coordinates plots by using [`ggparcoord()`](https://ggobi.github.io/ggally/reference/ggparcoord.html) of [**GGally**](https://r4va.netlify.app/chap15) package, and

-   plotting interactive parallel coordinates plots by using [**parallelPlot**](https://cran.r-project.org/web/packages/parallelPlot/) package.

## **15.2 Installing and Launching R Packages**

```{r}
pacman::p_load(GGally, parallelPlot, tidyverse)
```

## **15.3 Data Preparation**

In the code chunk below, `read_csv()` of **readr** package is used to import *WHData-2018.csv* into R and save it into a tibble data frame object called *wh*.

```{r}
wh <- read_csv("data/WHData-2018.csv")
```

## **15.4 Plotting Static Parallel Coordinates Plot**

In this section, you will learn how to plot static parallel coordinates plot by using [`ggparcoord()`](https://ggobi.github.io/ggally/reference/ggparcoord.html) of **GGally** package. Before getting started, it is a good practice to read the function description in detail.

### **15.4.1 Plotting a simple parallel coordinates**

Code chunk below shows a typical syntax used to plot a basic static parallel coordinates plot by using `ggparcoord()`.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12))
```

### **15.4.2 Plotting a parallel coordinates with boxplot**

The basic parallel coordinates failed to reveal any meaning understanding of the World Happiness measures. In this section, you will learn how to makeover the plot by using a collection of arguments provided by `ggparcoord()`.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Parallel Coordinates Plot of World Happines Variables")
```

### **15.4.3 Parallel coordinates with facet**

In the code chunk below, `facet_wrap()` of **ggplot2** is used to plot 10 small multiple parallel coordinates plots. Each plot represent one geographical region such as East Asia.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region)
```

### **15.4.4 Rotating x-axis text label**

To make the x-axis text label easy to read, let us rotate the labels by 30 degrees. We can rotate axis text labels using `theme()` function in ggplot2 as shown in the code chunk below.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30))
```

### **15.4.5 Adjusting the rotated x-axis text label**

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30, hjust=1))
```

## **15.5 Plotting Interactive Parallel Coordinates Plot: parallelPlot methods**

[**parallelPlot**](https://cran.r-project.org/web/packages/parallelPlot/) is an R package specially designed to plot a parallel coordinates plot by using ‘htmlwidgets’ package and [d3.js](https://d3js.org/).

### **15.5.1 The basic plot**

The code chunk below plot an interactive parallel coordinates plot by using `parallelPlot()`

```{r}
wh <- wh %>%
  select("Happiness score", c(7:12))
parallelPlot(wh,
             width = 320,
             height = 250)
```

### **15.5.2 Rotate axis label**

In the code chunk below, `rotateTitle` argument is used to avoid overlapping axis labels.

```{r}
parallelPlot(wh,
             rotateTitle = TRUE)
```

### **15.5.3 Changing the colour scheme**

We can change the default blue colour scheme by using `continousCS` argument as shown in the code chunl below.

```{r}
parallelPlot(wh,
             continuousCS = "YlOrRd",
             rotateTitle = TRUE)
```

### **15.5.4 Parallel coordinates plot with histogram**

In the code chunk below, `histoVisibility` argument is used to plot histogram along the axis of each variables.

```{r}
histoVisibility <- rep(TRUE, ncol(wh))
parallelPlot(wh,
             rotateTitle = TRUE,
             histoVisibility = histoVisibility)
```

# **16  Treemap Visualisation with R**

## **16.1 Overview**

In this hands-on exercise, we will gain hands-on experiences on designing treemap using appropriate R packages. The hands-on exercise consists of three main section. First, we will learn how to manipulate transaction data into a treemap strcuture by using selected functions provided in **dplyr** package. Then, we will learn how to plot static treemap by using **treemap** package. In the third section, we will learn how to design interactive treemap by using **d3treeR** package.

## **16.2 Installing and Launching R Packages**

```{r}
pacman::p_load(treemap, treemapify, tidyverse) 
```

## **16.3 Data Wrangling**

### **16.3.1 Importing the data set**

```{r}
realis2018 <- read_csv("data/realis2018.csv")
```

### **16.3.2 Data Wrangling and Manipulation**

-   group transaction records by *Project Name*, *Planning Region*, *Planning Area*, *Property Type* and *Type of Sale*, and

-   compute *Total Unit Sold*, *Total Area*, *Median Unit Price* and *Median Transacted Price* by applying appropriate summary statistics on *No. of Units*, *Area (sqm)*, *Unit Price (\$ psm)* and *Transacted Price (\$)* respectively.

-   Two key verbs of **dplyr** package, namely: *group_by()* and *summarize()* will be used to perform these steps.

### **16.3.3 Grouped summaries without the Pipe**

The code chunk below shows a typical two lines code approach to perform the steps.

```{r}
realis2018_grouped <- group_by(realis2018, `Project Name`,
                               `Planning Region`, `Planning Area`, 
                               `Property Type`, `Type of Sale`)
realis2018_summarised <- summarise(realis2018_grouped, 
                          `Total Unit Sold` = sum(`No. of Units`, na.rm = TRUE),
                          `Total Area` = sum(`Area (sqm)`, na.rm = TRUE),
                          `Median Unit Price ($ psm)` = median(`Unit Price ($ psm)`, na.rm = TRUE), 
                          `Median Transacted Price` = median(`Transacted Price ($)`, na.rm = TRUE))
```

### **16.3.4 Grouped summaries with the *pipe***

The code chunk below shows a more efficient way to tackle the same processes by using the *pipe*, %\>%:

```{r}
realis2018_summarised <- realis2018 %>% 
  group_by(`Project Name`,`Planning Region`, 
           `Planning Area`, `Property Type`, 
           `Type of Sale`) %>%
  summarise(`Total Unit Sold` = sum(`No. of Units`, na.rm = TRUE), 
            `Total Area` = sum(`Area (sqm)`, na.rm = TRUE),
            `Median Unit Price ($ psm)` = median(`Unit Price ($ psm)`, na.rm = TRUE),
            `Median Transacted Price` = median(`Transacted Price ($)`, na.rm = TRUE))
```

## **16.4 Designing Treemap with treemap Package**

**treemap** package is a R package specially designed to offer great flexibility in drawing treemaps. The core function, namely: *treemap()* offers at least 43 arguments. In this section, we will only explore the major arguments for designing elegent and yet truthful treemaps.

### **16.4.1 Designing a static treemap**

```{r}
realis2018_selected <- realis2018_summarised %>%
  filter(`Property Type` == "Condominium", `Type of Sale` == "Resale")
```

### **16.4.2 Using the basic arguments**

The code chunk below designed a treemap by using three core arguments of *treemap()*, namely: *index*, *vSize* and *vColor*.

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### **16.4.3 Working with *vColor* and *type* arguments**

In the code chunk below, *type* argument is define as value.

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type = "value",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### **16.4.5 The “value” type treemap**

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="value",
        palette="RdYlBu", 
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### **16.4.6 The “manual” type treemap**

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="RdYlBu", 
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

Single colour palette such as Blues is used:

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="Blues", 
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### **16.4.7 Treemap Layout**

*treemap()* supports two popular treemap layouts, namely: “squarified” and “pivotSize”. The default is “pivotSize”.

### **16.4.8 Working with *algorithm* argument**

The code chunk below plots a squarified treemap by changing the *algorithm* argument.

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="Blues", 
        algorithm = "squarified",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### **16.4.9 Using *sortID***

When “pivotSize” algorithm is used, *sortID* argument can be used to dertemine the order in which the rectangles are placed from top left to bottom right.

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="Blues", 
        algorithm = "pivotSize",
        sortID = "Median Transacted Price",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

## **16.5 Designing Treemap using treemapify Package**

### **16.5.1 Designing a basic treemap**

```{r}
ggplot(data=realis2018_selected, 
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`),
       layout = "scol",
       start = "bottomleft") + 
  geom_treemap() +
  scale_fill_gradient(low = "light blue", high = "blue")
```

### **16.5.2 Defining hierarchy**

Group by Planning Region

```{r}
ggplot(data=realis2018_selected, 
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`),
       start = "topleft") + 
  geom_treemap()
```

Group by Planning Area

```{r}
ggplot(data=realis2018_selected, 
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`,
           subgroup2 = `Planning Area`)) + 
  geom_treemap()
```

Adding boundary line

```{r}
ggplot(data=realis2018_selected, 
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`,
           subgroup2 = `Planning Area`)) + 
  geom_treemap() +
  geom_treemap_subgroup2_border(colour = "gray40",
                                size = 2) +
  geom_treemap_subgroup_border(colour = "gray20")
```

# 16.6.1 Designing An Interactive Treemap


```{r}
library(d3treeR)
```


```{r}
tm <- treemap(realis2018_summarised,
        index=c("Planning Region", "Planning Area"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="value",
        title="Private Residential Property Sold, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

d3tree() is used to build an interactive treemap.

```{r}
d3tree(tm,rootname = "Singapore" )
```
